The elevational structure of a mountain mutualism: modes and thresholds in the responses of bumble bees, wildflowers, and their interactions

### 2021-09-22 ms_1/ms_revision1.Rmd
analysis reboot

1. Biological responses to environmental gradients

Environmental gradients regulate the assembly of biological communities, and they have likely played a central role in generating and maintaining Earth's biological diversity [@Hutchinson1953-mo]. Patterns of species- and community-level responses to environmental gradients can shed light on the underlying mechanisms that govern community assembly and function [e.g. @Beals1969-in]. Such insight is important for anticipating and managing the effects of global ecological change that modify existing environmental gradients together with the species that inhabit them.

2. The particular case of mountain elevation gradients

Mountain slopes generate steep environmental gradients over short geographical distances, facilitating the assembly of biological communities with high species richness and turnover[@Rahbek2019-sw]. The response of species richness and turnover to mountain elevation gradients

*Species richness*

Species richness along mountain slopes tends to show either a mid-elevation peak or a monotonic decline [@Rahbek1995-oj; @Guo2013-tu], though responses can vary considerably across taxonomic groups [e.g. @Peters2016-hf].

*Species turnover*

In some cases, continuous elevation gradients can generate punctuated shifts in species composition. as has been demonstrated in some mountain plant communities [@Beals1969-in; @Hemp2006-bw]

Something about productivity ~ elevation

Maybe something about abundance ~ elevation

<!-- Vegetational Change Along Altitudinal Gradients; controversy regarding whether vegetation gradients are continuous or discrete, suggests something in between, consistent with the "threshold" that I see associated with the tree line. [@Beals1969-in] -->

<!-- *A "though see" citation:* The ecological and floristic importance of the forest limit in phytogeographical zonations has remained un- questioned,   even though its  value as an important ecological boundary has  rarely been verified. An interesting question is therefore  whether  the  presence or absence of birch  forests in Aurland is associated with any abrupt discontinuity in floristic richness or composition ... Our results from Aurland suggest that the forest limit is  not associated with  any major floristic change,  either in species turnover or in species richness, but is  associated with some  change in overall species composition. Hofgaard (1997) similarly found little difference in beta diversity or floristic composition 300 m above or below the tree line in the central Norwegian mountains.  -->

3. From species to interactions: less studied, but amenable to the same tools used at the species-level [see references in @Simanonok2014-ou]

While elevational patterns in gross summaries of ecological communities (e.g. species richness, net primary productivity) are valuable, potentially deeper insight can be gained by the more focused study of elevational patterns among guilds of species with known functional relationships, such as plants and herbivores or plants and pollinators.

An elevational study of plants and insect herbivores [@Rasmann2014-oc]

4. Mountain bumble bees and wildflowers
  A. Opportunity to study responses not just at the level of species, but at the level of interactions [@Simanonok2014-ou; @Burkle2016-tq].
  B. Significance of understanding the elevation structure of mountain BB-wildflower communities vis-a-vis climate change and meadow reforestation

5. What we did

Definitely need to discuss @Hoiss2012-sn and @Hoiss2012-kr and @Hoiss2015-gv

The response-effect framework might be useful [@Suding2008-bl].

<!-- First introduce the concept of ecological thresholds, nonlinear responses. Give some illustrative examples to show the general nature of the phenomenon.  -->

Concept paper on ecological thresholds [@Groffman2006-ch].

Threshold response of stream communities to *urbanization* gradient [@King2011-ry].

Threshold response of anurans to proximity to highways [@Eigenbrod2009-zl].

Nonlinear and threshold responses of nitrogen cycling to *aridity* gradient [@Wang2014-fh].

Threshold response of marine fungal community composition to *salinity* gradient [@Rojas-Jimenez2019-rm].

Threshold and nonlinear responses of marine benthic community to *disturbance* (sheer stress from waves, trawling, etc.) gradients [@Couce2020-hu].

Threshold response of bat communities to habitat loss gradient [@Muylaert2016-rj].

Mountains are unique ecological theaters in which global climate dynamics interact with local climate gradation to shape the distributions of species and biotic interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Peters2019-km; @Rafferty2020-aj]. The rate of recent climate change, however, is forcing species to adapt rapidly within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Elevational range shifts, however, are not merely abstract translations along the linear lapse of temperature generated by mountain slopes; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Thus, when species move along continuous, linear climate gradients, they can experience punctuated and nonlinear ecological transitions. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a linear elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir].

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms in the field of community ecology, and they have more recently become conservation priorities due to well-documented population declines and vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. Bumble bees are characteristically cold-adapted [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Upslope movement of bumble bee species in response to climate change has already been documented, and this is expected to cause the declines or extinction of some bumble bee species [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; Biella2017-ni; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. From a purely geometric perspective, populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer habitat shrinkage [@Dirnbock2011-jf], possibly accompanied by increasing genetic isolation [though see @Fijen2020-zg]. But the outcomes of elevational range shifts on bumble bees also hinge upon the current elevational distributions of bumble bees and their floral resources that will be acted upon by climate change. The upward advance of lowland bumble bee species, for example, may introduce novel competitive pressures on highland species and rewire the pollination networks in which they participate [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm], and elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their floral hosts. When range shifts are considered in this complex biotic context, it is especially important to understand nonlinearities or thresholds in species-, community-, and interaction-level responses to elevational climate gradients, since these could produce dramatic ecological consequences that would not be predicted by geometric considerations alone.

<!-- The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic process of community assembly under climate change, a phenomenon of both historical interest and prospective urgency.  -->
In the present study, set along a ~1400 m elevation gradient in the German Alps, we investigate elevational patterns in a community of bumble bees and wildflowers at three levels of biological organization. First, we ask how elevation shapes the abundance and richness of bumble bees and their floral resources. Then, we move from the level of individual species to that of community composition and explore the elevational structure of species turnover (i.e. β-diversity) in bumble bees and flowering plants. Finally, we advance to the level of species interactions, investigating the turnover of bumble- bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques designed to capture nonlinear responses to elevation, with a particular emphasis on the tree line ecotone as a key threshold that structures the biological communities of mountains slopes.

@@

The study was conducted 2010-2012 in Berchtesgaden National Park, located in the Northern Limestone Alps of southeast Germany (47.55°N, 12.92°E). The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 meters above sea level (m.s.l.) **(Figure 1)**. Fourteen of these pastures are extensively grazed by cattle or sheep, 3 are mowed for hay production, and 8 have been abandoned throughout the last century and are no longer subject to any human management. Study sites were classified as lying either above or below the tree line based on an examination of Google Earth imagery corroborated by field experience at our study sites. Eight sites were located above the tree line, which generally fell at an elevation of around 1500 m.s.l., consistent with the previous descriptions of the tree line in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn].

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory according to [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. Males are often found resting on flowers, particularly late in the year, so their presence on a flower cannot be interpreted reliably as a foraging visit. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^. Plant species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied: 153 in 2010, 212 in 2011, and 539 in 2012 **(Figure 2)**. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient.

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date. To weight floral abundance based on the foraging preferences of each bumble bee species in our network, we scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of *visited* genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We treated the cryptic *Bombus terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *Bombus barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.  *Bombus hypnorum*, *B. humilis*, and *B. lapidarius* were extremely rare in our samples, so we omitted them from abundance analyses but retained them for β-diversity analyses.

Eight plant species --- *Rubus idaeus*, *Rosa canina*, *Juniperus communis*, *Larix decidua*, *Salix* sp., *Caltha palustris*, *Pulmonaria officinalis*, and *Rheum barbarum* --- were recorded in visitation data but not recorded during floral surveying. Each accounted for no more than 4 visits in total over the three years of our study, and they were omitted from the analysis of floral abundance.

### Elevational distribution of bumble bees and floral resources

We analyzed the abundance of bumble bees and floral resources, respectively, using hierarchical generalized additive models (HGAMs) [@Wood2011-pg; @Pedersen2019-kf]. To focus on elevational patterns, we collapsed the temporal variation of abundance within sites into summary values. For bumble bees, we modeled *peak* abundance at each site as a smooth response to `elevation`, with a group-level smoother for each bumble bee `species`, `management` (grazing, mowing, or unmanaged) as a fixed effect, and `species:year` as a random intercept term (exploratory analysis indicated that species varied differently across years). For floral resources, we modeled *mean* flower cover, weighted per bumble bee species (i.e. the sum of *visited* genera), as a smooth response to `elevation`, with `management` as a fixed effect, and `year` and bumble bee `species` as random intercept terms. For the bumble bee model, we used the negative binomial distribution family (log link function), and for the floral model we used a Tweedie family distribution (log link function). Both models follow the "type I" model form described in @Pedersen2019-kf.

@@

For our analyses of species and interaction β-diversity, we pooled all samples (including across years) within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated species (separately for bumble bees and plants) and interaction β-diversity between all pairs of sites using the binary Jaccard index and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk]. Only plant species that were visited at least once by a bumble bee over the course of the study were included in the calculation of floral β-diversity.

<!-- and modeled each relationship using logistic matrix regression [@Goslee2009-ln]. Site management difference (i.e. same or different management category) was included as a covariate to account for potential management effects, and the difference in number of sampling dates between sites was included as covariate to control for the potential confounding effects of sampling frequency. -->

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of β-diversity, fitting each response with a binomial regression smooth. To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variable, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. We also included site management in each model by encoding management classes as a series of binary dummy variables, following @Ferrier2007-ji. To control for potential confounding effects of the number of sampling days per site, sampling days was included as a covariate. The importance of the variables in each model were assessed by a permutation test in which variable importance is defined as the percent loss of explained deviance when a given variable is permuted in isolation [@Fitzpatrick-2021aa].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa]. β-diversity was calculated with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Data and code to reproduce these analyses are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys.

<!-- Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. -->

@@

@@

## Elevational distribution of bumble bees and floral resources

HGAM analysis revealed that the response of peak abundance to elevation differed significantly (p ≤ 0.02) from a flat line for all bumble bee species **(Figure 2A)**. *Bombus mendax*, *B. monticola*, *B. mucidus*, *B. pyrenaeus*, and *B. soroeensis* exhibited a positive response to elevation and peaked in abundance above the tree line, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. In contrast, the abundance of *B. hortorum*, *B. pascorum*, *B. psithyrus*, and *B. terrestris/lucorum* declined with elevation. This effect was especially pronounced in *B. pascuorum*, likely because its exceptionally long nesting cycle is incompatible with the short growing season at high elevations [@Neumayer2009-gx]. *Bombus gerstaeckeri*, *B. jonellus*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance *at* the tree line. Our model explained 71% of total variance (adjusted R^2^ = 0.45). Neither mowing (p = 0.08) nor non-management (p = 0.95) had a significant effect on bumble bee abundance relative to the reference level of grazing management. The random effect of species:year was also nonsignificant (p = 0.52). See Appendix S1 for tabulated model output and diagnostics.

Mean floral resource abundance likewise responded significantly (p << 0.0001) to elevation for all bumble bee species **(Figure 2B)**. *Bombus jonellus*, *B. mendax*, *B. monticola*, *B. mucidus*, *B. soroeensis*, and *B. wurflenii* exhibited bimodal patterns of flora resource abundance, with peaks around 1000 and 1600 m.s.l, the latter peak corresponding closely to the location of the tree line. *Bombus pyrenaeus* exhibited a unimodal floral resource pattern peaking near 1600 m.s.l. The remaining species --- *B. gerstaeckeri*, *B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. psithyrus*, and *B. terrestris/lucorum* experienced linear or quadratic decline in floral resource abundance with increasing elevation. Floral resource abundance was low for all species at elevations above ~1600 m.s.l. Our model explained 67% of total variance (adjusted R^2^ = 0.41). Both lack of management (p << 0.0001) and mowing (p << 0.0001) had a significant positive effect on mean flower cover relative to the reference level of grazing management. The effect of year was also significant, with 2010 (p << 0.0001) and 2011 (p << 0.0001) exhibiting higher mean flower cover than 2012. See Appendix S1 for tabulated model output and diagnostics.

@@

GDM analysis revealed a strongly nonlinear response of floral and interaction β-diversity to elevation **(Figure 4A)**. Turnover was high from ~600-1000 m, representing the transition from valley to lower slopes, and then leveled off from ~1000-1500 m. At around 1500 m --- the elevation of the tree line --- turnover accelerated sharply and remained steep for the remainder of the elevation gradient. In contrast, bumble bee turnover was modest and exhibited no marked thresholds. Our GDMs explained 65% of total deviance in interaction β-diversity, 74% of floral β-diversity, and 27% of bumble bee β-diversity. The conditional effect of elevation was significant (p ~ 0) in all models, and its importance was 1-2 orders of magnitude higher that of sampling frequency and management **(Figure 4B)**. Nevertheless, the conditional effect of sampling frequency was significant (p = 0) in the interaction β-diversity model, and mowing management was significant (p = 0) in the floral β-diversity model.

@@

The core finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate complex and nonlinear biological responses in a mountain bumble- bee-wildflower community in terms of both abundance and β-diversity, and the latter both at the level of community composition and at the level of species interactions. These responses are punctuated by the tree line ecotone, which constitutes a discrete ecological threshold along the linear elevation gradient of a mountain slope.

>>>>

Environmental gradients regulate the assembly of biological communities, and they have likely played a central role in generating and maintaining Earth's biological diversity [@Hutchinson1953-mo]. Patterns of species- and community-level responses to environmental gradients can shed light on the underlying mechanisms that govern community assembly and function [e.g. @Beals1969-in]. In addition to its basic significance, such insight is important for anticipating and managing the effects of climate and land use change, which can modify existing environmental gradients and reorganize the biological communities that inhabit them.

Mountain slopes generate steep environmental gradients over short geographical distances, facilitating the assembly of biological communities with high species richness and turnover [@Rahbek2019-sw]. Perhaps the most salient environmental condition that changes along a mountain elevation gradient is air temperature, which declines according to a linear lapse rate of approximately -0.6°C per 100 m [@Rolland2003-my], but mountain slopes also generate variation in other conditions, including precipitation, atmospheric pressure, atmospheric turbulence, substrate characteristics, and exposure to solar radiation [@Korner1995-mh; @Hodkinson2005-af].

Ecologists have long viewed mountain slopes as natural laboratories for the study of community assembly, and particular attention has been given to elevational patterns of species richness and composition. Species richness along mountain slopes tends to exhibit either a mid-elevation peak or a monotonic decline [@Rahbek1995-oj; @Guo2013-tu], though the exact shape of these responses can vary considerably across taxonomic groups [e.g. @Peters2016-hf]. Species composition along mountain slopes, best studied in plant communities, can exhibit a range of patterns from smooth gradations to relatively abrupt transitions, with steeper slopes tending to exhibit more discrete patterns of species turnover [@Beals1969-in]. For example, many mountains have a well-defined tree line ecotone, where plant communities exhibit accelerated species turnover and declining diversity as multiple abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir; @Becker2007-cg; @Descombes2017-id; though see @Odland1999-vx].

<!-- In most cases, plant species composition along a mountain slope varies more or less continuously but with nonlinearities or thresholds of species turnover, suggesting an interaction of processes more complex than a simple response to the linear decline of temperature with elevation [@Beals1969-in].  -->

<!-- The peaking of richness at mid-elevations might reflect intermediate environmental conditions that favor species richness, but this pattern has also been attributed to the *mid-domain effect*, which predicts mid-elevation peaks as a geometric effect of bounded ranges independent of environmental conditions [@Colwell2000-qf]. Similarly, monotonic declines in richness with elevation could be explained by abiotic conditions becoming more stringent with elevation, but they are also consistent with a species-area effect driven by the geometric loss of surface area as a mountain's topography converges to a summit [@Rahbek1997-hj; though see @Elsen2015-gp].  -->

<!-- When discrete transitions in species composition occur along a continuous environmental gradient, they can be interpreted as points where the fitness inequality between competing species reverses, implying that interspecific competition is a primary mechanism of community assembly [@Hutchinson1953-mo].  -->

While elevational patterns in overall species richness and composition are informative, the functional significance of mountain slopes can be explored more deeply by focusing on the responses of interacting guilds of species. For example, studies of plants and insect herbivores along mountain elevation gradients have revealed a positive correlation between elevation and herbivore niche-breadth, suggesting an extension of the classical *latitude-niche-breadth hypothesis* to the analogous case of elevational gradients [@Rasmann2018-gl]. Similarly, plant-pollinator interactions along mountain slopes have been found to become less specialized as elevation increases, which may reflect a reciprocal relationship between the processes of competition and environmental filtering in community assembly [@Hoiss2015-gv; @Classen2020-nh]. A unifying theme of such studies is the understanding of *species interactions* as ecological entities *per se*, dependent upon but not reducible to the species-level richness and composition of the interacting guilds.

Bumble bees (Hymenoptera: *Bombus* spp.) and their floral hosts provide an excellent system in which to study elevational patterns at both the species- and interaction-level. Bumble bees are believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary, and their subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, they remain characteristically cold-adapted organisms [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes [@Williams1998-xi]. In these habitats, they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Goulson2010-lo]. As in the past, so now are mountain habitats critical theaters of ecological change for bumble bees. Warming climates are expected to cause habitat loss and potentially novel competitive pressures [@Marshall2020-pc; @Soroye2020-zb; though see @Guzman2021-my]. At the same time, socioeconomic drivers are causing the reforestation of pastures on mountain slopes and the intensification of agriculture in lowlands and valleys, both of which entail a loss of floral resources for bumble bees [@Tasser2002-om; @Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw].

In this paper, we present a study based on ~13,000 interactions between bumble bees (Hymenoptera: *Bombus*) and their floral hosts recorded along a 1400 m elevation gradient, complemented by corresponding surveys of floral abundance. Using this dataset, we investigate elevational responses of richness, abundance, and β-diversity (i.e. species turnover), both at the level of species (bumble bees and flowering plants, respectively) and at the level of interactions (bumble- bee-flower pairs). Our results reveal nonlinear and, in some cases, contrasting elevational patterns, underscoring the fact that mountain slopes are complex ecological systems where multiple processes interact to produce biological responses. These patterns shed light on the basic ecology of the mountain mutualism between bumble bees and wildflowers, and they suggest how this mutualism might be reshaped by ongoing processes of climate and land use change.

@@

The study was conducted from 2010-2012 in Berchtesgaden National Park, located in the Northern Limestone Alps of southeast Germany (47.55°N, 12.92°E). The landscape is composed of mountain pastures surrounded mainly by coniferous forest. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 meters above sea level (m.s.l.) **(Figure 1)**. Fourteen of these pastures are extensively grazed by cattle or sheep, 3 are mowed for hay production, and 8 have been abandoned throughout the last century and are no longer subject to any human management. Study sites were classified as lying either above or below the tree line based on an examination of Google Earth imagery corroborated by field experience at our study sites. Eight sites were located above the tree line, which generally fell at an elevation of around 1500 m.s.l., consistent with the previous descriptions of the tree line in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn].

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain when the air temperature was at least 6°C. During each visit, interactions between bumble bee and flowers were recorded during a 50-minute transect walk. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory according to [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. Males are often found resting on flowers, particularly late in the year, so their presence on a flower cannot necessarily be interpreted as a foraging visit. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^. Plant species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) and bumble bee observations varied across years **(Figure 2A-C)**. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient **(Figure 2D)**.

@@

We quantified bumble bee abundance as the total number of recorded floral visits per bumble bee species per site-date. We treated the cryptic *Bombus terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *Bombus barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.  *Bombus hypnorum*, *B. humilis*, and *B. lapidarius* were extremely rare in our samples, so we omitted them from abundance analyses but retained them for β-diversity analyses.

To focus our analysis on the subset if flowering plant species that are relevant to bumble bee foraging, we omitted from analysis all plant species that received no visits from bumble bees during the course of our study. Eight plant species --- *Rubus idaeus*, *Rosa canina*, *Juniperus communis*, *Larix decidua*, *Salix* sp., *Caltha palustris*, *Pulmonaria officinalis*, and *Rheum barbarum* --- were recorded as visited during interaction sampling but were missed during floral surveying. Each accounted for no more than 4 visits in total over the three years of our study, so they were also omitted from analysis.

### Elevational patterns of richness and abundance

We analyzed the richness and abundance of bumble bees, flowering plants, and their interactions, respectively, using hierarchical generalized additive models (HGAMs) [@Wood2011-pg; @Pedersen2019-kf]. To focus on elevational patterns, we collapsed within-year variation of richness and abundance at each site into annual summary values of cumulative richness and peak abundance. We then modeled these summary values, respectively, as smooth responses to elevation. Following the "GI" model form described by @Pedersen2019-kf, we specified both a global smoother and year-specific smoothers, with the latter understood as "random" deviations from the global effect. In each model, we also included `year` and `management` as fixed intercept effects, and we included `site` as a random intercept effect to avoid pseudoreplication due to repeated sampling of sites across years. In the richness models, we also added `number of transects` as a smoother to account for the effect of sampling intensity on observed richness. For all models, we used the quasipoisson (log-link) distribution family.

To explore species-level elevational responses in bumble bees, we fit a model with a species-wise elevation smoother and no global smoother, corresponding to the "I" model form of @Pedersen2019-kf. `Year` and `management` were included as fixed intercept effects and `site` and `species` were included as random intercept effects. For this model, we used the negative binomial (log link) distribution family because it yielded a better residual distribution than the quasipoisson family.

To facilitate clear interpretation with respect to our focal questions, the models above focus on elevational patterns of abundance by summarizing across seasonal time. For those interested in within-season abundance patterns and their interaction with elevation, more complex tensor-product GAMs are included in Appendix 1.

@@

For our analyses of species and interaction β-diversity, we pooled all samples by site irrespective of year. After pooling, we calculated β-diversity between all pairs of sites for bumble bees, flowering plants, and theirs interaction, respectively, using the binary Jaccard index. We then partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk].

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of β-diversity, fitting each response with a binomial regression smooth. To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and their (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variable, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. To control for potential confounding effects of sampling intensity, sampling days per site was included as a covariate. We also included the geographic distances between sites as a covariate to account for potential spatial autocorrelation. The importance of the variables in each model were assessed by a permutation test [@Fitzpatrick-2021aa].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017], `mgcViz` [@Fasiolo2018aa], and `gratia` [@Simpson2021-aa]. β-diversity was calculated with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Data and code to reproduce these analyses are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys.

@@

**Check to make sure this summary is still true**

@@

## Elevational patterns of richness and abundance

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then approximately leveled off **(Figure 3A)**. Neither mowing nor abandonment had a significant effect on abundance relative to the reference management level of grazing (p > 0.05), but richness was significantly lower in 2012 than in the reference year of 2010 (p = 0.04) **(Table 1)**. In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 3B)**. Management had no significant effect on floral richness (P > 0.1), but floral richness was significantly lower in 2011 (p < 0.001) and 2012 (p = 0.03) than in 2010. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 3C)**. Management did not have a significant effect on interaction richness (p > 0.1), but interaction richness was significantly higher in 2011 (p < 0.05) and lower in 2012 (p < 0.001) relative to the reference level of 2010.

Cumulative bumble bee abundance did not respond significantly to elevation (p > 0.1) or management (p > 0.1), but bumble bee abundance was significantly higher in 2011 than in the reference year of 2010 (p = 0.03) **(Figure 4A; Table 2)**. Floral abundance, however, peaked around 1000 and exhibited an approximately quadratic decline with increasing elevation (p = 0.003). Floral abundance was higher in abandoned meadows than in grazed meadows (p = 0.002) and lower in 2012 than in 2010 (p = 0.02).

At the species level, bumble bees exhibited variable and sometimes divergent responses to elevation **(Figure 5)**. *Bombus pyrenaeus*, *B. mendax*, *B. monticola*, and *B. mucidus* exhibited approximately linear increases with elevation (p < 0.05). In contrast, *Bombus psithyrus* and *B. pascuorum* exhibited approximately linear decreases with elevation (p < 0.001). *Bombus pratorum* was unique in exhibiting a strongly unimodal response to elevation peaking around 1600 m (p < 0.001). The remaining species did not respond significantly to elevation (p > 0.05).

@@

Bumble bees, flora, and their interactions exhibited marked differences in their rates of β-diversity in response to elevation. **(Figure 4)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral β-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 3)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee β-diversity, 73% of floral β-diversity, and 71% of interactions β-diversity.

@@

## Big picture

The core finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate complex and nonlinear biological responses in a mountain bumble- bee-wildflower community in terms of richness, abundance and β-diversity, and at the level of both species and interactions.

The positive response of bumble bee richness to elevation sets bumble bees apart from most other taxa [@Rahbek1995-oj; @Guo2013-tu; @Peters2016-hf], underscoring their exceptional adaptedness to mountain environments. On higher mountains, the increase of bumble bee richness with elevation eventually gives way to a unimodal response, but peak richness is often found above 3000 m [e.g. @Williams1991-zu; @Streinzer2019-bz], ~1000 m above the highest peaks in the Berchtesgadener Alps. Thus, in our study system, the positive response of bumble bee richness to elevation encompasses the entire mountain slope, in agreement with the findings of @Minachilis2020-xd along the comparable elevation gradient (300---2700 m) of Mt. Olympus. Importantly, the positive response of bumble bee richness stands in sharp contrast to the monotonic decline of overall bee richness with elevation in the Berchtesgadener Alps [@Hoiss2012-sn], highlighting the special importance of bumble bees as pollinators where few other bees persist.
